The state learner -- a super learner for right-censored data

### 2025-05-21 manuscript/manuscript-one.tex
use cases section draft

% %%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
% %
% % This is a general template file for the LaTeX package SVJour3
% % for Springer journals.          Springer Heidelberg 2010/09/16
% %
% % Copy it to a new file with a new name and use it as the basis
% % for your article. Delete % signs as needed.
% %
% % This template includes a few options for different layouts and
% % content for various journals. Please consult a previous issue of
% % your journal as needed.
% %
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% %
% % First comes an example EPS file -- just ignore it and
% % proceed on the \documentclass line
% % your LaTeX will extract the file if required
% \begin{filecontents*}{example.eps}
% %!PS-Adobe-3.0 EPSF-3.0
% %%BoundingBox: 19 19 221 221
% %%CreationDate: Mon Sep 29 1997
% %%Creator: programmed by hand (JK)
% %%EndComments
% gsave
% newpath
%   20 20 moveto
%   20 220 lineto
%   220 220 lineto
%   220 20 lineto
% closepath
% 2 setlinewidth
% gsave
%   .4 setgray fill
% grestore
% stroke
% grestore
% \end{filecontents*}
% %
% \RequirePackage{fix-cm}
% %
% %\documentclass{svjour3}                     % onecolumn (standard format)
% %\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
% \documentclass[smallextended]{svjour3}       % onecolumn (second format)
% %\documentclass[twocolumn]{svjour3}          % twocolumn
% %
% \smartqed  % flush right qed marks, e.g. at end of proof
% %
% \usepackage{graphicx}
% %
% % \usepackage{mathptmx}      % use Times fonts if available on your TeX system
% %
% % insert here the call for the packages your document requires
% \usepackage{dsfont, amssymb,soul,xcolor, enumitem, amsmath,verbatim}
% \usepackage[authoryear]{natbib}
% \definecolor{linkcolor}{rgb}{0, 0, 0.54}
% \usepackage[colorlinks,allcolors=linkcolor]{hyperref}
% %
% % please place your own definitions here and don't use \def but
% \DeclareMathOperator{\E}{\mathbb{E}} % expectation
% \newcommand{\Z}{\mathbb{Z}}
% \newcommand{\R}{\mathbb{R}}
% \newcommand{\N}{\mathbb{N}}
% \newcommand{\C}{\mathbb{C}}
% \renewcommand{\S}{\mathbb{S}}
% \newcommand{\blank}{\makebox[1ex]{\textbf{$\cdot$}}}
% \newcommand\independent{\protect\mathpalette{\protect\independenT}{\perp}}
% \def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}}
% \renewcommand{\phi}{\varphi}
% \renewcommand{\epsilon}{\varepsilon}
% \newcommand*\diff{\mathop{}\!\mathrm{d}}
% \newcommand{\weakly}{\rightsquigarrow}
% \newcommand\smallO{\textit{o}}
% \newcommand\bigO{\textit{O}}
% \newcommand{\midd}{\; \middle|\;}
% \newcommand{\1}{\mathds{1}}
% \usepackage{ifthen} %% Empirical process with default argument
% \newcommand{\G}[2][n]{
% {\ensuremath{\mathbb{G}_{#1}}{\left[#2\right]}}
% }
% \DeclareMathOperator*{\argmin}{\arg\!\min}
% \DeclareMathOperator*{\argmax}{\arg\!\max}
% \newcommand{\empmeas}{\ensuremath{\mathbb{P}_n}} % empirical measure
% \newcommand{\data}{\ensuremath{\mathcal{D}}}

% \begin{document}

% \title{The state learner -- a super learner for right-censored
%   data}

% \author{Anders Munch \& Thomas A. Gerds}

\maketitle

\begin{abstract}
  In survival analysis, prediction models are needed as
  stand-alone tools and in applications of causal inference to
  estimate nuisance parameters. The super learner is a machine
  learning algorithm which combines a library of prediction
  models into a meta learner based on cross-validated loss. In
  right-censored data, the choice of the loss function and the
  estimation of the expected loss need careful consideration.
  We introduce the state learner, a new super learner for
  survival analysis, which simultaneously evaluates libraries
  of prediction models for the event of interest and the
  censoring distribution. The state learner places no
  restrictions on the algorithms that can be included in the
  library, works in the presence of competing risks, and does
  not require a single pre-specified estimator of the
  conditional censoring distribution. We establish an oracle
  inequality for the state learner and investigate its
  performance through numerical experiments. We illustrate the
  application of the state learner with prostate cancer data.

  % \keywords{Competing risks, cross-validation, loss based estimation,
  %   right-censored data, super learner}
% \PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

@@

\section{The concept of super learning}
\label{sec:super-learning}

@@

censoring each quadruple
$(\Lambda_{1}, \Lambda_{2}, \Gamma, H)$ characterises a
distribution \(P\in\mathcal P\), c.f.\

@@

      & =

@@

The state learner requires three libraries of learners (c.f., Section \ref{sec:super-learning}),
\(\mathcal{A}_1\), \( \mathcal{A}_2 \), and \( \mathcal{B} \), where
\(\mathcal{A}_1\) and \( \mathcal{A}_2\) contain learners for the
conditional cause-specific cumulative hazard functions \(\Lambda_1\)
and \( \Lambda_2\), respectively, and \(\mathcal{B}\) contains
learners for the conditional cumulative hazard function of the
censoring distribution. % We further

@@

approximately equally sized subsets (see Section \ref{sec:super-learning}), each learner

@@

\( \mathcal{F}(\mathcal{A}_1, \mathcal{A}_2, \mathcal{B}) \) is
evaluated using the cross-validated loss,

@@

The parameter \( F \) is not of much interest...

@@

...

Risk predictions can then be obtained by combining
\( \Lambda_1 \) and $\Lambda_2$ from the highest ranked tuple
using a well-known formula \citep{benichou1990estimates,
  ozenne2017riskregression}.

@@

% BibTeX users please use one of
% \bibliographystyle{spbasic}      % basic style, author-year citations
%\bibliographystyle{spmpsci}      % mathematics and physical sciences
%\bibliographystyle{spphys}       % APS-like style for physics

\bibliographystyle{abbrvnat}
\bibliography{bib.bib}   % name your BibTeX data base

>>>>

\begin{abstract}
  In survival analysis, prediction models are needed as
  stand-alone tools and in applications of causal inference to
  estimate nuisance parameters. The super learner is a machine
  learning algorithm which combines a library of prediction
  models into a meta learner based on cross-validated loss. In
  right-censored data, the choice of the loss function and the
  estimation of the expected loss need careful consideration.
  We introduce the state learner, a new super learner for
  survival analysis, which simultaneously evaluates libraries
  of prediction models for the event of interest and the
  censoring distribution. The state learner places no
  restrictions on the algorithms that can be included in the
  library, works in the presence of competing risks, and does
  not require a single pre-specified estimator of the
  conditional censoring distribution. We establish an oracle
  inequality for the state learner and investigate its
  performance through numerical experiments. We illustrate the
  application of the state learner with prostate cancer data.
\end{abstract}

\textit{\textbf{Keywords:} Competing risks, cross-validation,
  loss based estimation, right-censored data, super learner}

@@

\section{Super learning}
\label{sec:super-learning}

@@

censoring each tuple $(\Lambda_{1}, \Lambda_{2}, \Gamma, H)$
characterises a distribution \(P\in\mathcal P\), c.f.\

@@

      =

@@

The state learner requires three libraries of learners:
\(\mathcal{A}_1\), \( \mathcal{A}_2 \), and \( \mathcal{B} \),
where \(\mathcal{A}_1\) and \( \mathcal{A}_2\) contain
learners for the conditional cause-specific cumulative hazard
functions \(\Lambda_1\) and \( \Lambda_2\), respectively, and
\(\mathcal{B}\) contains learners for the conditional
cumulative hazard function of the censoring
distribution. % We further

@@

approximately equally sized subsets (c.f., Section
\ref{sec:super-learning}), each learner

@@

\( \mathcal{F}(\mathcal{A}_1, \mathcal{A}_2, \mathcal{B}) \)
is evaluated using the cross-validated loss,

@@

The state learner estimates the parameter \( F \) which is a
feature of the observed right-censored data distribution
\( P \). In particular, \( F \) depends on the censoring
distribution and is typically not of direct interest in
itself. In the following, we describe how an estimate of
\( F \), as provided by the state learner, can be used to
estimate more relevant parameters of interest, such as
survival probabilities.

@@

We here provide explicit formulas for obtaining event-free
survival probabilities and risk predictions given the
state-occupation function \( F \) defined in
equation~\eqref{eq:F-def}. These formula holds under the
assumption of conditional independent censoring and
positivity, as introduced in Section~\ref{sec:framework}.

By equations~(\ref{eq:lambdaj}) and~(\ref{eq:gamma}) and the
definition of \( F \), we have
\begin{equation}
  \label{eq:7}
  % \Gamma(t , x) = \int_0^t \frac{F(\diff s, -1, x )}{F(s-,
  % 0, x )}, \quad \text{and} \quad
  \Lambda_j(t , x) 
  = \int_0^t  \frac{F(\diff s, j, x )}{F(s-, 0, x )},
  \quad j \in \{1,2\}.
\end{equation}
Equation~\eqref{eq:surv-def} provides a formula for obtaining
the event-free survival probabilities from the cause-specific
cumulative hazard functions. Similarly, cause-specific risk
predictions can be obtained from \( \Lambda_1 \) and
$\Lambda_2$ using the formula
\citep[e.g.,][]{benichou1990estimates,
  ozenne2017riskregression},
\begin{equation}
  \label{eq:cs-risk-def}
  Q(T \leq t, D = j \mid X=x) =
  \int_0^t \exp\left\{-\Lambda_{1}(u \mid x)-\Lambda_{2}(u
    \mid x)\right\}  \Lambda_j(\diff u \mid x),
  \quad j \in \{1,2\}.
\end{equation}
Hence, given the state learner's estimate of \( F \) we can
use equation~\eqref{eq:7} to obtain estimates of the
cause-specific cumulative hazard functions $\Lambda_j$, which
can in turn be used to obtain estimates of the event-free
survival probabilites and cause-specific risks through
equations~\eqref{eq:surv-def} and~\eqref{eq:cs-risk-def}.

In Section~\ref{sec:super-learner-simple} we suggested to
implement the state learner by building a library using
learners of the cause-specific cumulative hazard functions,
$\Lambda_j$, and the cumulative hazard function for censoring,
$\Gamma$. With this implementation we can directly input the
highest ranked tuple of cause-specific hazard functions
$(\Lambda_1, \Lambda_2)$ provided by the state learner as
input to equations~\eqref{eq:surv-def}
and~\eqref{eq:cs-risk-def}.

@@

\bibliography{bib.bib}
